---
title: "Article 12 (Salary Schedule) Data"
author: "Melissa Anderson, Ph.D."
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#install libraries
library(readxl)
library(tidyverse)
library(here)
library(readxl)
library(kableExtra)
library(ggplot2)
library(scales)
library(dplyr)
library(formattable)
library(tinytex)

#load in inflation data
inflation_coefficients <- read_excel(here("data", "Inflation Coefficients.xlsx"))

#load in CFT Budget Data
budget_data <- read_excel(here("data", "Budget Data.xlsx"))
salary_data_2324 <- read_excel(here("data", "CFT Time Series (Clean).xlsx"), sheet = "2324")
salary_data_2223 <- read_excel(here("data", "CFT Time Series (Clean).xlsx"), sheet = "2223")
salary_data_2122 <- read_excel(here("data", "CFT Time Series (Clean).xlsx"), sheet = "2122")
salary_data_2021 <- read_excel(here("data", "CFT Time Series (Clean).xlsx"), sheet = "2021")
salary_data_1920 <- read_excel(here("data", "CFT Time Series (Clean).xlsx"), sheet = "1920")
salary_data_1819 <- read_excel(here("data", "CFT Time Series (Clean).xlsx"), sheet = "1819")
salary_data_1718 <- read_excel(here("data", "CFT Time Series (Clean).xlsx"), sheet = "1718")

#Merge Data into a single data frame
by = join_by(District)
all_data <- full_join(salary_data_2324,salary_data_2223, by)
all_data <- full_join(all_data,salary_data_2122, by)
all_data <- full_join(all_data,salary_data_2021, by)
all_data <- full_join(all_data,salary_data_1920, by)
all_data <- full_join(all_data,salary_data_1819, by)
all_data <- full_join(all_data,salary_data_1718, by)

#Select Salary Data and Make Long Version
all_data_salary <- select(all_data, District, Adjustment, contains("Salary"))
all_data_salary_long <- pivot_longer(data = all_data_salary, 
                              cols =  "2324_Salary_MA_0_1":"1718_Salary_Doctorate_Max",
                              names_to = "category",
                              values_to = "salary")
#Create Year/Classification Columns
all_data_salary_long <- all_data_salary_long %>% 
  mutate(category = str_replace(category, pattern = "2324_", replacement = "2023-"))
all_data_salary_long <- all_data_salary_long %>% 
  mutate(category = str_replace(category, pattern = "2223_", replacement = "2022-"))
all_data_salary_long <- all_data_salary_long %>% 
  mutate(category = str_replace(category, pattern = "2122_", replacement = "2021-"))
all_data_salary_long <- all_data_salary_long %>% 
  mutate(category = str_replace(category, pattern = "2021_", replacement = "2020-"))
all_data_salary_long <- all_data_salary_long %>% 
  mutate(category = str_replace(category, pattern = "1920_", replacement = "2019-"))
all_data_salary_long <- all_data_salary_long %>% 
  mutate(category = str_replace(category, pattern = "1819_", replacement = "2018-"))
all_data_salary_long <- all_data_salary_long %>% 
  mutate(category = str_replace(category, pattern = "1718_", replacement = "2017-"))

all_data_salary_long <- all_data_salary_long %>% 
  tidyr::separate(category, into = c("year", "classification"), sep = "-")
all_data_salary_long <- all_data_salary_long %>% 
  mutate(year = as.numeric(year) )

#add in inflation coefficients
by <- join_by(year == Year)
all_data_salary_long <- left_join(all_data_salary_long,inflation_coefficients,by)

```

# PCC full-time salaries are low compared to neighboring districts

```{r echo=FALSE}

#Identify Peer Districts
set.seed(32)

data_classification <- select(all_data, District, Zip, Adjustment, "2324_FTES", "2324_General_Fund_Revenue")
data_classification <- data_classification %>% 
  filter(District != "California Online (Calbright)")
data_classification <- data_classification %>% 
  mutate(Zip = as.numeric(Zip))

data_classification_scaled <- scale(data_classification[,2])

kmi <- kmeans(data_classification_scaled, 3)
clusters <- data.frame(data_classification$District, kmi$cluster)


#aggregate district/classification data
all_data_salary_long$salary <- currency(all_data_salary_long$salary, digits = 2L)
X2023_data_salary_long <- all_data_salary_long %>% 
  filter(year == 2023)
X2023_data_salary_long <- X2023_data_salary_long %>% 
  mutate(salary_COL = salary/Adjustment)

X2023_data_salary_long <- X2023_data_salary_long[order(X2023_data_salary_long$salary_COL),]

X2023_data_salary_long_plot <- X2023_data_salary_long %>%
  group_by(District) %>%
  summarize(mean_salary = mean(salary),
            mean_salary_COL = mean(salary_COL))

X2023_data_salary_long_plot <- X2023_data_salary_long_plot[order(X2023_data_salary_long_plot$mean_salary_COL, decreasing = TRUE),]

#add in district geographical classification
by <- join_by(District == data_classification.District)
X2023_data_salary_long_plot <- left_join(X2023_data_salary_long_plot,clusters,by)

X2023_data_salary_long_plot <- X2023_data_salary_long_plot %>% 
  filter(kmi.cluster == 2)

PCC_Salary_2023 <- as.numeric(X2023_data_salary_long_plot[8, "mean_salary"])

X2023_data_salary_long_plot <- X2023_data_salary_long_plot %>% 
  mutate(percent_compare = percent(mean_salary/PCC_Salary_2023))

X2023_data_salary_long_plot <- select(X2023_data_salary_long_plot, District, mean_salary, percent_compare)

X2023_data_salary_long_plot[,1:3] %>% kbl(caption = "Table 1: Area CCC District Representative Salaries During the 2023-204 Academic Year", centering = TRUE,col.names = c("District", "Aggregate Salary", "Percent Comparison to PCC Salary"), position = "left") %>%
  kable_classic(full_width = T) %>%
  row_spec(8, bold = T, color = "white", background = "#9c1521")

  


```

* Aggregate Salary is composite value made by averaging values for the five representative salary steps found in the CFT Salary Report.  *(Data Source: https://www.cft.org/faculty-salary-comparisons)* 



# PCC full-time salaries have not significantly increased relative to inflation

```{r echo=FALSE}
PCC_data_salary_long <- all_data_salary_long %>% 
  filter(District == "Pasadena Area")


PCC_data_salary_long <- PCC_data_salary_long %>% 
  mutate(
    salary_inf = salary * `Inflation Coeff`
  )

PCC_data_salary_long_plot <- PCC_data_salary_long %>%
  group_by(year) %>%
  summarize(mean_salary_inf = mean(salary_inf))

initial_salary <- PCC_data_salary_long_plot$mean_salary_inf[1]
salary_range <-  PCC_data_salary_long_plot$mean_salary_inf[1]
  #max(PCC_data_salary_long_plot$mean_salary_inf)-min(PCC_data_salary_long_plot$mean_salary_inf)        #sd(PCC_data_salary_long_plot$mean_salary_inf)

PCC_data_salary_long_plot <- PCC_data_salary_long_plot %>% 
  mutate(
    rel_mean_salary_inf = (mean_salary_inf - initial_salary)/salary_range
  )


ggplot(data = PCC_data_salary_long_plot, aes(x = year, y = rel_mean_salary_inf)) +
  geom_line(color = "#9c1521") +
    scale_y_continuous("Relative Change",
                       labels = scales::percent) +
  theme_minimal() + 
  theme(legend.position="bottom")
```

**Figure 1:** Relative Changes in Aggregate Salary of Full-Time PCC Faculty since 2017. Values are in 2024 dollars to remove the effect of inflation. Aggregate Salary is a composite value made by averaging values for the five representative salary steps found in the CFT Salary Report. **Note that when inflation is not taken into account, even the maximum increase in PCC salaries (3%) was much less than the difference between PCC Salaries most neighboring districts (1-14%) and subsequent inflation eroded a significant portion of those gains.** (Data Source: https://www.cft.org/faculty-salary-comparisons)



# The District has seen a significant increase in income relative to expenses and total compensation
```{r budget_time_plot, echo=FALSE}
#load in CFT Budget Data
budget_data <- read_excel(here("data", "Budget Data.xlsx"))

#create data frame for plot
by <- join_by(Year)
budget_data_inf <- left_join(budget_data,inflation_coefficients,by)
budget_data_inf <- budget_data_inf %>% 
  mutate(
    actual_total_income_inf = `Actual Total Income` * `Inflation Coeff`,
    actual_total_expenses_inf = `Actual Total Expenses` * `Inflation Coeff`,
    actual_total_compensation_inf = `Actual Total Compensation` * `Inflation Coeff`
  )

actuals_data_inf <- budget_data_inf %>% select(Year, `Actual Total Income`, `Actual Total Expenses`, `Actual Total Compensation`, actual_total_income_inf, actual_total_expenses_inf, actual_total_compensation_inf)

actuals_data_inf_long <- pivot_longer(data = actuals_data_inf, 
                                    cols = actual_total_income_inf:actual_total_compensation_inf,
                                    names_to = "category",
                                    values_to = "amount")

ggplot(data = subset(actuals_data_inf_long, !is.na(amount)), aes(x = Year, y = amount, color = category)) +
  geom_line() +
    scale_y_continuous("Amounts",
    labels = scales::label_dollar()
    ) +
  theme_minimal()

```

**Figure 2:** Total Income, Expenses and Compensation Expenses over time. Values have been adjusted to 2024 dollars to remove the effect of inflation. *(Data Source: https://pasadena.edu/business-administrative-services/fiscal-services/budget-forecast-analysis.php)* 


# Extra funds could be used for a significant increase to total compensation

```{r how_much_is_left, echo=FALSE}

a = 10
pa = a/100
b = 1 + pa
f = 2

reserve_inf <- budget_data_inf %>% select(Year, actual_total_income_inf, actual_total_expenses_inf, actual_total_compensation_inf)

reserve_inf <- reserve_inf %>% 
  filter(!is.na(actual_total_income_inf))

reserve_inf <- reserve_inf %>% 
  mutate(
    total_reserve = actual_total_income_inf - actual_total_expenses_inf,
    p_reserve = actual_total_income_inf/actual_total_expenses_inf - 1,
    excess = (actual_total_income_inf - actual_total_expenses_inf - pa*actual_total_compensation_inf)/actual_total_expenses_inf,
    excess2 = (actual_total_income_inf - actual_total_expenses_inf - pa*actual_total_compensation_inf/f)/actual_total_expenses_inf
  )

  
reserve_diff_inf <- reserve_inf %>% select(Year, total_reserve, p_reserve, excess, excess2)
reserve_diff_inf$total_reserve <- currency(reserve_diff_inf$total_reserve, digits = 2L)
reserve_diff_inf$p_reserve <- percent(reserve_diff_inf$p_reserve,1)
reserve_diff_inf$excess <- percent(reserve_diff_inf$excess, 1)
reserve_diff_inf$excess2 <- percent(reserve_diff_inf$excess2, 1)


reserve_diff_inf %>%
  kbl(caption = "Table 1: Amounts of District Income Left Over After Expenses are Deducted (2024 Dollars)", centering = TRUE,col.names = c("Year", "Total Left", "% of Total Expenses", "Minus Raise A*", "Minus Raise B*"), position = "left") %>%
  kable_classic(full_width = T)

```
* Percentage of Total Expenses (i.e. "Reserve") Left after a (A) `r a`%  or (B) `r round(a/f,1)`%  Increase in Total Compensation. *(Data Source: https://pasadena.edu/business-administrative-services/fiscal-services/budget-forecast-analysis.php)* 

```{r eval=FALSE, include=FALSE}
"# Useful Links
* *Update on community college reserves.* https://lao.ca.gov/Publications/Report/4323. (January 27, 2021)
* *Faculty salary comparison studies. CFT – a Union of Educators and Classified Professionals.* https://www.cft.org/faculty-salary-comparisons. 
* *Budget / Forecast & Analysis - Business and Administrative Services - Pasadena City College.* https://pasadena.edu/business-administrative-services/fiscal-services/budget-forecast-analysis.php."
```
